\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-31 paper.qmd
Add first scratch of the discussion

- Summarise contributions
- Explain where it is expected or not expected to work, eg higher dimensional relationships
- Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views
- Predicting new observations in $k$-D
- Extending layouts beyond $k$-D, when 2D is clearly inadequate.
- Diagnostic app to explore differences in distances
- What might be useful enhancements

>>>>

<!-- - Summarise contributions -->
<!-- - Explain where it is expected or not expected to work, eg higher dimensional relationships -->
<!-- - Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views (need to add)-->
<!-- - Predicting new observations in $k$-D -->
<!-- - Extending layouts beyond $k$-D, when 2D is clearly inadequate. -->
<!-- - Diagnostic app to explore differences in distances (need to add)--> 
<!-- - What might be useful enhancements -->

We have developed an algorithm to evaluate the most useful NLDR method and (hyper-)parameter choices for creating a reasonable \gD{} layout of high-dimensional data. Our objective is to fit a model for the \gD{} layout that preserves the relationships between neighboring points and turns it into a high-dimensional wireframe, which can be overlaid on the data and visualized using a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

The S-curve example demonstrates how the model accurately fits the points. Viewing the wireframe with a tour reveals how well the model represents the underlying geometry of the S-curve, highlighting its ability to preserve both local and global structures in high-dimensional space.

As demonstrated in the simulation case study, all observed NLDR methods preserve the global structure, while tSNE is the only method that preserves the local structure. Our approach helps in determining this and illustrates how different methods have specific quirks.

Even though three distant well-separated clusters have shown in the UMAP layout of **pbmc** suggested by @chen2023, the model evident that the clusters are close to each other in high dimension. Also, viewing the model in the data space shows data patches which shows the data are not uniformly distributed within the clusters and non-linear structures within clusters. These summaries can't be observed in UMAP layout. Therefore, model in high dimension is useful to obtain some quirks like these.

On the other hand, the error or unexplained variance is important in evaluating how well the model fit the data. By observing error for different number of bins, tSNE with perplexity 30 selects as a reasonable representation for **pbmc**. Connecting closest clusters by line segments in the fitted model provides evidence of preserving the neighborhood.

We have shown in the **digit: 1**, how the model accurately represents the non-linear structure. Also, provides some unobserved information in \gD{}. The twisted pattern which squeezed the structure in some projections and long line segments which detect anomalies.

Predicting new observations in \kD{} is particularly valuable due to the limitations of some NLDR methods, like tSNE, which don't provide a straightforward method for prediction. As a result, our approach offers a solution that capable of generating predicted \kD{} embedding regardless of the NLDR method employed, effectively addressing this functional gap.

The future enhancements involve introducing approaches to bind the data and indicate line segments beyond \gD{}. This would help in creating a more accurate representation of the data when \gD{} layout is inadequate.

## Supplementary Materials

Code, and data for reproducing this paper are available at [https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm).

### 2024-05-31 paper.tex
Add first scratch of the discussion

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  Human behaviour, the desire to have more certainty, and a tendency to
  prefer the well-separated views
\item
  Predicting new observations in \(k\)-D
\item
  Extending layouts beyond \(k\)-D, when 2D is clearly inadequate.
\item
  Diagnostic app to explore differences in distances
\item
  What might be useful enhancements
\end{itemize}

>>>>

We have developed an algorithm to evaluate the most useful NLDR method
and (hyper-)parameter choices for creating a reasonable \gD{} layout of
high-dimensional data. Our objective is to fit a model for the \gD{}
layout that preserves the relationships between neighboring points and
turns it into a high-dimensional wireframe, which can be overlaid on the
data and visualized using a tour. This approach is defined as
\emph{model-in-data-space}. Viewing a model in the data space is an
ideal way to examine the fit.

The S-curve example demonstrates how the model accurately fits the
points. Viewing the wireframe with a tour reveals how well the model
represents the underlying geometry of the S-curve, highlighting its
ability to preserve both local and global structures in high-dimensional
space.

As demonstrated in the simulation case study, all observed NLDR methods
preserve the global structure, while tSNE is the only method that
preserves the local structure. Our approach helps in determining this
and illustrates how different methods have specific quirks.

Even though three distant well-separated clusters have shown in the UMAP
layout of \textbf{pbmc} suggested by \citet{chen2023}, the model evident
that the clusters are close to each other in high dimension. Also,
viewing the model in the data space shows data patches which shows the
data are not uniformly distributed within the clusters and non-linear
structures within clusters. These summaries can't be observed in UMAP
layout. Therefore, model in high dimension is useful to obtain some
quirks like these.

On the other hand, the error or unexplained variance is important in
evaluating how well the model fit the data. By observing error for
different number of bins, tSNE with perplexity 30 selects as a
reasonable representation for \textbf{pbmc}. Connecting closest clusters
by line segments in the fitted model provides evidence of preserving the
neighborhood.

We have shown in the \textbf{digit: 1}, how the model accurately
represents the non-linear structure. Also, provides some unobserved
information in \gD{}. The twisted pattern which squeezed the structure
in some projections and long line segments which detect anomalies.

Predicting new observations in \kD{} is particularly valuable due to the
limitations of some NLDR methods, like tSNE, which don't provide a
straightforward method for prediction. As a result, our approach offers
a solution that capable of generating predicted \kD{} embedding
regardless of the NLDR method employed, effectively addressing this
functional gap.

The future enhancements involve introducing approaches to bind the data
and indicate line segments beyond \gD{}. This would help in creating a
more accurate representation of the data when \gD{} layout is
inadequate.

\section{Supplementary Materials}\label{supplementary-materials}

Code, and data for reproducing this paper are available at
\url{https://github.com/JayaniLakshika/paper-nldr-vis-algorithm}.
